### 2020-02-02 nlp_models.tex
updates

In this section, we examine quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pre-trained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrics~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger $\alpha$ PL exponents.
Based on HT-SR Theory (in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus that they are substantially \emph{under}-trained.
%
More generally, these NLP models are quite different than the CV models we considered in Section~\ref{sxn:cv}, both in terms of their architecture (e.g., wheter there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our metrics.

\michael{WHERE: Norm-base metrics perform reasonably well on good NLP models, but they have problems with less well-trained models.
PL-based metrics (the Log $\alpha$-Norm metric, $\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$, and the Weighted Alpha metric, $\hat\alpha =\alpha\log\lambda_{max} $) do even better.
We highlight how to use the theory to identify poorly trained models where the empirical norm metrics will perform poorly.}

\michael{WHERE: We show that this not only works for CV, so we look at NLP.
We now look in detail at the alpha metric, the spectral norm, and the new norm.
We compare good and bad models, and a series of increasingly big models
trained on big data sets.  We show why alpha works but spectral norm is not enough.
The layer Spectral Norm behaves unexpectedly, while $\alpha$-based metrics can give insight into how well trained a model is.}

\michael{WHERE: Say large $\alpha$ values don't mean much.}

\michael{WHERE: Highlight difference between $\alpha$ and $\hat{\alpha}$.}

>>>>

In this section, we examine quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pre-trained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrics~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger $\alpha$ PL exponents.
Based on HT-SR Theory, 
%(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to successfully capture many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties, and 
%, both in terms of their architecture (e.g., wheter there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
thus, they provide a very different test for our empirical quality metrics.

Where Norm-base metrics perform reasonably well on well-trained NLP models, they behave anomalously poorly-trained models.
Indeed, in ``bad models'', the weight matrics may display rank collapse, decreased Frobeinus mass, or unsually small Spectral norms, 
which may be mis-interpreted as ``smaller is better.''
In contast, PL-based metrics (the Log $\alpha$-Norm metric, $\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$, and the Weighted Alpha metric, $\hat\alpha =\alpha\log\lambda_{max} $) display consistant behavior even on poorly trained models, and we can use these metrics to help indentify
where the archiectures need repair, and/or more and/or better data is needed.

%We now look in detail at the alpha metric, the spectral norm, and the new norm.
%We compare good and bad models, and a series of increasingly big models
%trained on big data sets.  We show why alpha works but spectral norm is not enough.
%The layer Spectral Norm behaves unexpectedly, while $\alpha$-based metrics can give insight into how well trained a model is.

\michael{WHERE: Say large $\alpha$ values don't mean much.}
We do note, however, that many NLP models, such as the GPT and BERT, may have weight matrices with unusually large PL exponents
($\alpha>6$), which indicates these matrices may be \emph{under}-correlated (i.e over-paramaterized), although the truncated PL fit
itself may not be very reliable because the MLE estimator is unreliable in this range.  Phenomenologically, if we examine
the ESD visually, we can usually describe these $\mathbf{W}$ as in the \emph{Bulk-Decay} or \emph{Bulk-plus-Spikes} phase\cite{MM}.
We have conjectured previously that very well-trained DNNs would not have many \emph{outlier} $\alpha>6$, 
and improved versions of GPT (shown below, and BERT (not shown) confirm this.

\michael{WHERE: Highlight difference between $\alpha$ and $\hat{\alpha}$.}
To avoid confusion, let us clarify the differences between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated power law (PL), parameterized by 2 values:
the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.  (Technically, we also need the minimum
eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.)
The PL exponent $\alpha$ measures of the 
amount of correlation in a DNN layer weight matrix $\mathbf{W}$.  It is valid for $\lambda<\lambda_{max}$, and
it scale invariant (does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.)
The Weighted-Alpha metric, $\hat{\alpha}=\alpha\log\lambda_{max}$, approximates the Log $\alpha$ (Shatten) Norm 
$(\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha})$ (derived using statistical mechanics and RMT \nred{(in progress)}, 
and lets us compute a balanced, weighted average log PL norm metrix for the entire DNN.  
The $\hat{\alpha}$ metric also behaves like an imrpoved, weighted log average Spectral Norm, and may track this metric in some cases.
